--- 
title: "A Tour of the Generalized Lotka-Volterra Model"
site: bookdown::bookdown_site
author: "Stefano Allesina"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
description: 'Lectures for the ICTP-SAIFR/IFT-UNESP *School on Community Ecology: from patterns to principles* '
documentclass: book
link-citations: yes
bibliography: book.bib
biblio-style: apalike
---

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

```{r knitr1, echo=FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  comment   = "#",
  results   = "hold",
  message = FALSE,
  warning = FALSE,
  # collapse  = TRUE,
  fig.align = "center")
```

# Introduction

These lectures were prepared for the ICTP-SAIFR/IFT-UNESP "School on Community Ecology: from patterns to principles", held in São Paulo, Brazil, January 20-25, 2020.

The four 90-minute lectures provide a semi-rigorous introduction to the Generalized Lotka-Volterra model, along with its applications to community ecology. Throughout the text, I have included code that can be used to simulate the processes that are introduced from a mathematical point of view. 

The emphasis is squarely on models for large communities, composed of $n \gg 1$ species. In particular, I often refer to "random" communities, i.e., dynamical systems with random parameters. This allows me to derive interesting results that are meant to describe the "typical" or "expected" community.

At the end of the lecture, I refer to my own work in the area, highlighting some of the strongest contributions from my laboratory at the University of Chicago. 

## Prerequisites

The lectures are fairly self-contained, but some working knowledge of calculus, dynamical systems, linear algebra, game theory would be beneficial.

The code is written in `R`, and meant to run using a fairly recent version of `R` along with the packages `tidyverse`, `deSolve`, `MASS`, `lpSolve`... that need to be installed.

## Sources

These notes were prepared by summarizing and interpolating a variety of sources, including @hadeler2017topics, @hofbauer1998evolutionary, and @baigent2016lotka.

